\documentclass{scrartcl}

\input{../../shared.tex}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{lemma}{Lemma}[section]
\usepackage{enumerate}

\begin{document}


\section*{Problem set: Gradient descent and Fixed points}%



\paragraph{Exercise (i)} (1P)  Let $f_1, f_2, \dots, f_m$ be smooth with parameters $L_1, L_2, \dots L_m$. Show that the function $f:= \sum_{i=1}^{m}f_i$ is smooth with parameter $\sum_{i=1}^{m}L_i$.

\paragraph{Exercise (ii)} (1P)  Let $f$ be smooth with parameter $L$ and $A$ a matrix. Show that $f\circ A$ is smooth with parameter $L \Vert A \Vert^2$.




\subsection*{Computing Fixed Points}%

Gradient descent turns up in a surprising number of situations which apriori have nothing to do with optimization.
In this exercise we will see how computing the fixed point of functions can be seen as a form of gradient descent.
Suppose that we have a $1$-Lipschitz continuous function $g : \R \to \R$ such that we want to solve for
\begin{equation}
  g(x) = x .
\end{equation}
A simple strategy for finding such a fixed point is to run the following algorithm: starting from an arbitary $x_0$,
we iteratively set
\begin{equation}
  \label{fpi}
  x_{k+1} = g(x_k) .
\end{equation}

\paragraph{Exercise (iii)} (3P) Enter the missing code snippets in the jupyter notebook. Partial credit will be awarded.

We will try solve for $x$ starting from $x_0 = 1$ in the following two equations:
\begin{equation}
  \label{log}
  x = \log(1 + x), \quad \text{and} \quad x = \log(2 + x).
\end{equation}
What difference do you observe in the rate of convergence between the two problems? Let’s understand why this happens:

\paragraph{Exercise (iv)} (3P) Theoretical fixed point questions.
\begin{itemize}
  \item We want to re-write the update~\eqref{fpi} as a step of gradient descent. To do this, we need to find a function $f$
        such that the gradient descent update is identical to~\eqref{fpi}:
        \begin{equation}
          x_{k+1} = x_k - \alpha f'(x_k) = g(x_k) .
        \end{equation}
        Derive such a function $f$.
  \item Give sufficient conditions on $g$ to ensure convergence of procedure~\eqref{fpi}. What $\alpha$ would you need to pick?
        Hint: We know that gradient descent on f with fixed step-size converges if $f$ is convex and smooth. What
        does this mean in terms of $g$?
  \item What condition does $g$ need to satisfy to ensure linear convergence? Are these satisfied for the problems in~\eqref{log}
\end{itemize}


\end{document}
